---
title: "Lab-08b: Intro to Regular Expressions"
subtitle: "Stat 133"
format: 
  html:
    toc: true
    number-sections: true
    theme: zephyr
embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)  # ecosystem of data science packages 
library(rvest)      # for web scraping
library(plotly)     # for web-interactive plots
```

::: {.callout-note icon="false"}
## General Instructions

-   Use the template file `lab08b-template.qmd` to answer the questions.

-   Rename this file as `lab08b-first-last.qmd`, where `first` and `last` are your first and last names (e.g. `lab08b-gaston-sanchez.qmd`).

-   Make sure the YAML header in your `qmd` files includes `embed-resources: true`.

-   Submit both your qmd and HTML files to the corresponding assignment submission in bCourses.

-   Please note that if you submit the incorrect files you will receive no credit.
:::

::: callout-important
We are assuming that you've reviewed the practice material associated to lab-08: `practice-regex2.html`.

File in course website:

<https://stat133.berkeley.edu/fall-2024/practice/practice-regex2.html>
:::

------------------------------------------------------------------------

# Data: Men's 1500-meters Freestyle

You will be working with data for one of the world records in swimming, namely the *Men's long course 1500-meters freestyle world record progression*, available in wikipedia:

<https://en.wikipedia.org/wiki/World_record_progression_1500_metres_freestyle>

The data is part of an HTML table from the above wikipedia page.

## Data Scraping

One way to scrape the data is with some `"rvest"` functions:

tbl

```{r}
# wikipedia's url
# (assemble url so that code fits within screen)
wiki = "https://en.wikipedia.org/wiki/"
swim1500 = "World_record_progression_1500_metres_freestyle"
url = paste0(wiki, swim1500)

# scrape HTML tables (into a list)
tables = read_html(url) |>
  html_table()
  
# desired table is the first one
tbl = tables[[1]]
tbl
```

------------------------------------------------------------------------

# Data Cleaning and Timeline

The goal is to clean the scraped data table in order to produce a simpler and cleaner table via string manipulations with regex. More specifically, the *clean* tibble will have the following columns:

-   `time` (in minutes, as numeric)
-   `name` (name of athlete)
-   `month` (name of month)
-   `day` (number of day, as numeric)
-   `year` (number of year, as numeric)
-   `date` (date yyyy-mm-dd, as Date)

In addition, you will have to create a timeline graphic to visualize the progression of world records.

## Tip 1) Time Conversion to Minutes

As you can tell, the `Time` values are originally expressed as a mix of minutes and seconds. For example, the first record set by Henry Taylor back in Jul-25-1908 was **22:48.4** that is: 22 minutes, and 48.4 seconds.

To clean the data and obtain a column `time`, we need to convert those times into minutes:

$$
22 + 48.4/60 = `r 22 + 48.4/60`
$$

## Tip 2) Conversion to Date type

The values in column `Date` are given in non-standard format. Again, the date of the first record in the table is expressed as `"Jul 25, 1908"`.

To convert this string as an actual `"Date"` type, we can use `as.Date()`

```{r}
date1 = as.Date("Jul 25, 1908", "%B %d, %Y")
date1
```

In case you are curious, the specified date format has an interesting structure: `"%B %d, %Y"`. What do these symbols mean?

-   `%B` is a placeholder to indicate the *capitalized* name of the month

-   `%d` is a placeholder for the day of the month

-   `%Y` is a placeholder to indicate a four-digit year value

In turn, the output returned by `as.Date()` has a standard Date format: `"yyyy-mm-dd"`

------------------------------------------------------------------------

# Your Solutions

```{r}
# your code (feel free to use multiple code chunks)
time_column <- pull(tbl, "Time")
name_column <- pull(tbl, "Name")
date_column <- pull(tbl, "Date")

formatted_time_column <- str_split(time_column, ":")
minute_time_column <- rep(0, length(formatted_time_column))
for (i in 1:length(formatted_time_column)) {
  minute_time_column[i] = as.numeric(formatted_time_column[[i]][1]) + as.numeric(formatted_time_column[[i]][2]) / 60
}
formatted_date_column <- as.Date(date_column, "%B %d, %Y")

formatted_data = tibble(time = minute_time_column, 
                        name = name_column, 
                        month = month(formatted_date_column),
                        day = day(formatted_date_column),
                        year = year(formatted_date_column),
                        date = formatted_date_column
                      )


plot <- ggplot(data = formatted_data, aes(x = date, y = time)) +
  geom_point() +
  geom_line() +
  labs(title = "World record for Men’s 1500-meters Freestyle",
       x = "date",
       y = "time (minutes)") +
  theme_minimal()

ggplotly(plot)
```
