---
title: "HW2) Lists and Matrices"
subtitle: "Stat 133, Fall 2024"
author: "Wenyi Shi"
format: 
  html:
    toc: true
    number-sections: true
    theme: default
embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

**Stat 133**

> "Fiat Lux"

Concepts in **_computing_** with data

Some ~~stroke~~ text

This is a [hyperlink](hyperlink)




::: {.callout-note icon=false}

## General Instructions

-   Write your code and content in a `qmd` (quarto markdown) file. You can use
    the template file `hw2-template.qmd` provided in bCourses.

-   Name this file as `hw2-first-last.qmd`, where `first` and `last`
    are your first and last names (e.g. `hw2-gaston-sanchez.qmd`).

-   Make sure the YAML header in your `qmd` files includes `embed-resources: true`

-   Submit both your qmd and HTML files to the corresponding assignment
    submission in bCourses.

-   Please note that submitting only one of the files will result in an
    automatic 10% deduction.

-   Also, if you submit the incorrect files you will receive no credit.

:::


-----


# Creating a List

Refer to the instructions file `hw2-problems.html`, specifically the refer to
the list `index_fund`.

Write code to create the list `index`, and display this object (i.e. print it). 
Notice that `stats` is a named double vector, `stocks` is a list with 5 named 
elements, `characteristics` is a factor, and `performance` is a $3 \times 3$ 
matrix with row and column names.

```{r}
# function to compute letter grade "P" or "NP"
letter_grade <- function(x = 100) {
  if (!is.numeric(x)) {
    x <- as.numeric(x)
  }
  grade <- rep("P", length(x))
  grade[x < 70] <- "NP"
  grade[is.na(x)] <- NA
  return(grade)
}

tokelvin <- function(x) {
  return(273.15 + (5/9) * (x - 32))
}

x = sample(1:10, size = 1)
x

count = 0
for (i in c(2,4,6,8,10)) {
  for (j in 1:20) {
    for (k in 1:10) {
      # call some function we have defined elsewhere
      count = count + 1
    }
  }
}
count

print("===========")

# your code
stats = c(122.1800, 0.1681, 0.1815, 0.0300)
names(stats) = c("NAV", "YTD", "Stdev", "Expense")

stocks = list(
  "MSFT" = "Microsoft Corp",
  "APPL" = "Apple Inc",
  "NVDA" = "Nvidia Corp",
  "AMZN" = "Amazon Com Inc",
  "GOOG" = "Alphabet Inc"
)

characteristics_vec <- c("broad-based", "low-fee", "long-term")
characteristics <- factor(characteristics_vec)

performance = rbind(c(23.22, 7.92, 14.04), c(23.21, 7.95, 14.05), c(23.20, 9.90, 14.04))
rownames(performance) = c("return", "price", "benchmark")
colnames(performance) = c("1y", "3y", "5y")

index_fund = list(
  "description" = "Total U.S. Stock Market",
  "stats" = stats,
  "stocks" = stocks,
  "characteristics" = characteristics,
  "performance" = performance
)

index_fund
```



<br>

-----


# Harry Potter Characters

Refer to the Harry Potter data set described in the instructions file 
`hw2-problems.html`.


## Creating Vectors

Create vectors for the columns in the data table displayed above, 
according to the following data types. If there are missing values, encode them 
as `NA`. 

- `name`: character vector (name of individual)

- `born`: integer vector (years)

- `height`: real (i.e. double) vector (cm)

- `weight`: real (i.e. double) vector (lbs)

- `wand`: integer vector (length of wand, in inches)


```{r}
# your code
result <- 3

for (j in seq(from = 1, to = 18, by = 2)) {
  if (j %% 3 == 0) {
    result <- result - j
  } else {
    result <- result + j
  }
}


last = c("Stark", "Snow", "Lannister", "Targaryen", "Greyjoy")
first = c("Arya", "Jon", "Tyrion", "Daenerys", "Yara")
gpa = c(3,3,4,3,4)
got = data.frame(last, first, gpa)
got






name <- c("Harry", "Ronald", "Hermione", "Albus", "Rubeus")
born <- c(1980L, 1980L, 1979L, 1881L, 1928L)
height <- c(165, 175, 165, 180, 350)
weight <- c(130, 152, 118, 168, 289)
wand <- c(11L, 12L, 10L, 15L, 16L)

name
```


<br>

-----


# Matrix FIVE

## Creating a Matrix

Use the vectors created in the preceding part to create a matrix `FIVE`. 
This matrix should have dimensions: 5 rows, and 4 columns (born, height, 
weight, and wand). Also, give names to both the rows and columns of 
`FIVE` according to figure 1; and display this matrix.

```{r}
# your code
FIVE = matrix(c(born, height, weight, wand), nrow = 5, ncol = 4)
rownames(FIVE) = name
colnames(FIVE) = c("born", "height", "weight", "wand")
FIVE
```


## Subscripting (aka Indexing, Slicing) a Matrix

Refer to the matrix `FIVE`.

a) Using `FIVE`, write a command that gives you the first 3 rows, and 
columns `height` and `weight`.

```{r}
# your code
FIVE[1:3, c("height", "weight")]
```

<br>

b) Using `FIVE`, write a command that gives you the following output 
(see below): age values arranged in decreasing order. _Hint_: `sort()` is
your friend.

```
   Harry   Ronald Hermione   Rubeus    Albus 
    1980     1980     1979     1928     1881
```

```{r}
# your code
sort(FIVE[,"born"], decreasing = TRUE)
```

<br>

c) Using `FIVE`, write a command that gives you the name of the oldest 
individual. _Hint_: `which.min()` is your friend.

```{r}
# your code
name[which.min(FIVE[,"born"])]
```

<br>

d) Using `FIVE`, write a command that gives you the number of individuals that
have a wand length greater than or equal to 13. _Hint_: the function `sum()` is 
your friend.

```{r}
# your code
sum(FIVE[,"wand"] >= 13)
```

<br>

e) Using `FIVE`, write a command that arranges its content by `weight`
in decreasing order. _Hint_: the function `order()` is your friend.

```{r}
# your code
FIVE[order(FIVE[,"weight"], decreasing = TRUE), ]
```

<br>

f) Add a new row to `FIVE` with the data of Bellatrix given below. _Hint_: the
`rbind()` function is your friend. Display the updated `FIVE` matrix. To 
clarify: `FIVE` should have 6 rows (Bellatrix's data in row 6), and it should 
also include Bellatrix's name in the row names. 

- name: Bellatrix
- born: 1951
- height: 157
- weight: NA
- wand (length): 13

```{r}
# your code
Bellatrix <- c(1951L, 157, NA, 13L)
FIVE <- rbind(FIVE, Bellatrix)
FIVE
```

<br>

g) Add a column `ratio` to `FIVE`. The values of `ratio` should be the 
ratio of height over weight (i.e. height divided by weight). _Hint_: the 
`cbind()` function is your friend. Display the updated `FIVE` matrix. To 
clarify: `FIVE` should have 6 rows and 5 columns. 

```{r}
# your code
ratio <- FIVE[,"height"] / FIVE[,"weight"]
FIVE <- cbind(FIVE, ratio)
FIVE
```


<br>


-----


# List of Harry Potter Characters

Create a list `five` containing the vectors created in problem 2.
This list should have five elements (one vector per element). Make sure each 
element of the list has the corresponding name so that you can refer to them
by invoking either `five$name` or `five[["name"]]`. 
Display your `five` list.

```{r}
# your code
five <- list(
  "name" = name,
  "born" = born,
  "height" = height,
  "weight" = weight,
  "wand" = wand
)
five
```


## Subscripting (aka Indexing, Slicing) a List

__Refer to the list `FIVE`__. Write R commands---displaying the output---that 
answer the following questions. Answer each part in its individual code chunk.


a) What is the name of the individual with smallest wand length?
_Hint_: the function `which.min()` is your friend.

```{r}
# your code
five$name[which.min(five$wand)]
```

<br>

b) How many individuals have an unknown `born` value? _Hint_: the function 
`is.na()` is your friend.

```{r}
# your code
sum(is.na(five$born))
```

<br>

c) What is the median `born` year (ignore missing values)?

```{r}
# your code
median(five$born, na.rm = TRUE)
```

<br>

d) What is the height of the individual with the smallest weight value?
_Hint_: the function `which.min()` is your friend.

```{r}
# your code
five$height[which.min(five$weight)]
```

<br>

e) How many individuals have height values between 180 and 360 meters (inclusive)?
_Hint_: the function `sum()` is your friend.

```{r}
# your code
sum(five$height >= 180 & five$height <= 360)
```

<br>

f) Sort the names in `five` arranged---from smallest to largest---by their 
weight/height ("weight over height") values. _Hint_: the function `order()` is 
your friend.

```{r}
# your code
five$name[order(five$weight / five$height, decreasing = FALSE)]
```

<br>


-----


## Rubric {.unnumbered .unlisted}

- 😞 &nbsp; **None (0 pts)**: No submission; or terribly wrong; or incorrect files.

- 😟 &nbsp; **Deficient (1 pt)**: Displays minimal effort. Fairly incomplete. Code
    is badly written and/or has various major issues.

- 🙁 &nbsp; **Needs improvement (2 pts)**: Displays some effort. Doesn't complete all
    components. Code is poorly written and/or has various minor issues, possibly 
    including a major issue or mistake.

- 😐 &nbsp; **Fair (3 pts)**: Decent effort. Hits all the elements. Possibly
    a couple of minor mistakes or some non-minor issue. Code written succinctly, 
    but it could be better documented, and/or formatted.
    
- 🙂 &nbsp; **Good (4 pts)**: Solid effort. Hits all the elements. Possibly a 
    minor mistake. Fairly easy to follow (both the code and the output).

- 😀 &nbsp; **Excellent (5 pts)**: Finished all components of the assignment
    correctly (without any issues, bugs, mistakes). Code written succinctly, 
    well documented, and easy to follow (both the code and the output).

